---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# VicmapR

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![Codecov test coverage](https://codecov.io/gh/JustinCally/VicmapR/branch/master/graph/badge.svg)](https://codecov.io/gh/JustinCally/VicmapR?branch=master)
<!-- badges: end -->

The goal of VicmapR is to provide functions to easily access Victorin Government spatial data through their WFS (Web Feature Service). The package is currently in an early development stage.

## Installation

You can install the the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("JustinCally/VicmapR")
```
## Example

### Searching for data

```{r example}
library(VicmapR)

listLayers(pattern = stringr::regex("trees", ignore_case = T))
```

### Reading in data

As of VicmapR version `0.1.0` data is read in using a lazy evaluation method with the convenience of pipe operators (`%>%`). A lot of the methods and code have already been written for a similar package ([bcdata](https://github.com/bcgov/bcdata)) that downloads data from the British Columbia WFS catalogue. Using a similar approach to [bcdata](https://github.com/bcgov/bcdata), VicmapR allows users to construct a WFS query in a step-wise format. In doing so a query is reserved until `collect()` is used on the `vicmap_promise`. The example below shows an extensive example of how the to easily read in spatial data:

```{r query_example}
# Read in an example shape to restrict our query to using geometric filtering
melbourne <- sf::st_read(system.file("shapes/melbourne.geojson", package="VicmapR"), quiet = T)

# Obtain a promise of what data will be returned for a given layer
vicmap_query(layer = "datavic:VMHYDRO_WATERCOURSE_DRAIN")

# Build a more specific query and collect the results
vicmap_query(layer = "datavic:VMHYDRO_WATERCOURSE_DRAIN") %>% # layer to query
  filter(HIERARCHY == "L") %>% # simple filter for a column
  filter(INTERSECTS(melbourne)) %>% # more advanced geometric filter
  select(HIERARCHY, PFI) %>% 
  collect()
```

