---
title: "How to Query Vicmap Data"
output: rmarkdown::html_vignette
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{How to Query Vicmap Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  echo = TRUE,
  message = FALSE,
  comment = "#>"
)
```

```{r setup}
library(VicmapR)
library(sf)
library(leaflet)
```

# How to query data

## Searching for data  

In order to begin a query of the WFS server a spatial layer must be selected. To know which layers are available use the `listLayers()` function, which will return ~ 690 layers to choose from. 

```{r}
available_layers <- listLayers()

head(available_layers, 10)
```


## Vicmap promise

VicmapR introduces a new class called `vicmap_promise`, which is an extension to the `httr::url` class. Essentially this object is how the vicmap query is stored before data is collected. That is to say `vicmap_promise` is essentially a promise of what data will be retrieved. 

In order to generate a new promise the `vicmap_query` function can be used to select the layer. The promise prints a sample of the data (max = 6 rows) as well as the dimensions (nrow and ncol).  

```{r query}
# query the watercourse layer
vicmap_query(layer = "datavic:VMHYDRO_WATERCOURSE_DRAIN")
```

## Adding arguments to the query  

The `vicmap_promise` object can be easily added to through piping in of additional functions (e.g. `head()`, `filter()` and `select()`). 

The resulting query can be displayed using the `show_query()` function, which will list the WFS parameters. 

```{r query_arguments}
vicmap_query(layer = "datavic:VMHYDRO_WATERCOURSE_DRAIN") %>%
  head(50) %>% #return only 50 rows
  filter(HIERARCHY == "L") %>% # filter the column 'HIERACHY' to values of 'L'
  select(HIERARCHY, PFI) %>% # select columns 'HIERARCHY' and 'PFI'
  show_query()
```

In order to return a spatial data.frame object (`sf`) `collect()` must be used.

```{r collect_query}
watercourse_data <- vicmap_query(layer = "datavic:VMHYDRO_WATERCOURSE_DRAIN") %>%
  head(50) %>% #return only 50 rows
  filter(HIERARCHY == "L") %>% # filter the column 'HIERACHY' to values of 'L'
  select(HIERARCHY, PFI) %>% # select columns 'HIERARCHY' and 'PFI'
  collect()

str(watercourse_data)
```


## Geometric filters

VicmapR translates numerous geometric filter functions available in the Victorian Government's WFS Geoserver supports numerous [geometric filters](https://docs.geoserver.org/stable/en/user/tutorials/cql/cql_tutorial.html#geometric-filters):  

+ `EQUALS`   
+ `DISJOINT`   
+ `INTERSECTS`   
+ `TOUCHES`   
+ `CROSSES`   
+ `WITHIN`   
+ `CONTAINS` 
+ `OVERLAPS`   
+ `DWITHIN`   
+ `BEYOND`  
+ `BBOX`  

These filters can be used within the `filter()` function by providing them an object of class `sf/sfc/sfg/bbox`. Below is a leaflet map with the melbourne rail network being read in with the use of three different types of filter functions: `INTERSECTS()`, `BBOX()` and `DWITHIN()`.

```{r filter_spatial, warning = FALSE, message=FALSE, out.width="100%"}
#### Return objects that intersect melbourne ####

# Read in an example shape to restrict our query to using geometric filtering
melbourne <- sf::st_read(system.file("shapes/melbourne.geojson", package="VicmapR"), quiet = T)
# Return data that intersects melbourne
rail_intersects <- vicmap_query(layer = "datavic:VMTRANS_TR_RAIL") %>% # layer to query
  filter(INTERSECTS(melbourne)) %>% # more advanced geometric filter
  collect()

str(rail_intersects)

rail_bbox <- vicmap_query(layer = "datavic:VMTRANS_TR_RAIL") %>%
  filter(BBOX(sf::st_bbox(melbourne))) %>%
  collect()

rail_dwithin <- vicmap_query(layer = "datavic:VMTRANS_TR_RAIL") %>%
  filter(DWITHIN(melbourne %>% sf::st_centroid(), distance = 10000, units = "meters")) %>%
  collect()

leaflet(width = "100%") %>%
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons(data = melbourne, color = "grey", group = "Melbourne polygon") %>%
  addPolygons(data = sf::st_bbox(melbourne) %>% st_as_sfc(), color = "black", group = "Melbourne bbox") %>%
  addPolylines(data = rail_intersects, color = "Red", group = "INTERSECTS") %>% 
  addPolylines(data = rail_bbox, color = "Blue", group = "BBOX") %>%
  addPolylines(data = rail_dwithin, color = "Green", group = "DWITHIN") %>%
  addLayersControl(baseGroups = c("Melbourne polygon", "Melbourne bbox"), 
                   overlayGroups = c("INTERSECTS", "BBOX", "DWITHIN")) %>%
  hideGroup(c("BBOX", "DWITHIN"))
```


