---
title: "How to Query Vicmap Data"
output: rmarkdown::html_vignette
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{How to Query Vicmap Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  echo = TRUE,
  comment = "#>"
)
```

```{r setup}
library(VicmapR)
library(sf)
```

# How to query data

## Searching for data  

In order to begin a query of the WFS server a spatial layer must be selected. To know which layers are available use the `listLayers()` function, which will return ~ 690 layers to choose from. 

```{r}
available_layers <- listLayers()

head(available_layers, 10)
```


## Vicmap promise

VicmapR introduces a new class called `vicmap_promise`, which is an extension to the `httr::url` class. Essentially this object is how the vicmap query is stored before data is collected. That is to say `vicmap_promise` is essentially a promise of what data will be retrieved. 

In order to generate a new promise the `vicmap_query` function can be used to select the layer. The promise prints a sample of the data (max = 6 rows) as well as the dimensions (nrow and ncol).  

```{r query}
# query the watercourse layer
vicmap_query(layer = "datavic:VMHYDRO_WATERCOURSE_DRAIN")
```

## Adding arguments to the query  

The `vicmap_promise` object can be easily added to through piping in of additional functions (e.g. `head()`, `filter()` and `select()`). 

The resulting query can be displayed using the `show_query()` function, which will list the WFS parameters. 

```{r query_arguments}
vicmap_query(layer = "datavic:VMHYDRO_WATERCOURSE_DRAIN") %>%
  head(50) %>% #return only 50 rows
  filter(HIERARCHY == "L") %>% # filter the column 'HIERACHY' to values of 'L'
  select(HIERARCHY, PFI) %>% # select columns 'HIERARCHY' and 'PFI'
  show_query()
```

In order to return a spatial data.frame object (`sf`) `collect()` must be used.

```{r collect_query}
watercourse_data <- vicmap_query(layer = "datavic:VMHYDRO_WATERCOURSE_DRAIN") %>%
  head(50) %>% #return only 50 rows
  filter(HIERARCHY == "L") %>% # filter the column 'HIERACHY' to values of 'L'
  select(HIERARCHY, PFI) %>% # select columns 'HIERARCHY' and 'PFI'
  collect()

str(watercourse_data)
```


## Geometric filters

Vicmap translates numerous geometric filter functions available in the Victorian Government's WFS Geoserver supports numerous [geometric filters](https://docs.geoserver.org/stable/en/user/tutorials/cql/cql_tutorial.html#geometric-filters):  

+ `EQUALS`   
+ `DISJOINT`   
+ `INTERSECTS`   
+ `TOUCHES`   
+ `CROSSES`   
+ `WITHIN`   
+ `CONTAINS`   
+ `OVERLAPS`   
+ `RELATE`   
+ `DWITHIN`   
+ `BEYOND`  
+ `BBOX`  

These filters can be used within the `filter()` function by providing them an object of class `sf/sfc/sfg/bbox`. 

```{r filter_spatial}
#### Return objects that intersect melbourne ####

# Read in an example shape to restrict our query to using geometric filtering
melbourne <- sf::st_read(system.file("shapes/melbourne.geojson", package="VicmapR"), quiet = T)
# Reset crs to avoid GDAL issues
st_crs(melbourne$geometry) <- 4326
# Return data that intersects melbourne
watercourse_intersects <- vicmap_query(layer = "datavic:VMHYDRO_WATERCOURSE_DRAIN") %>% # layer to query
  filter(INTERSECTS(melbourne)) %>% # more advanced geometric filter
  collect()

watercourse_bbox <- vicmap_query(layer = "datavic:VMHYDRO_WATERCOURSE_DRAIN") %>%
  filter(BBOX(sf::st_bbox(melbourne))) %>%
  collect()

watercourse_dwithin <- vicmap_query(layer = "datavic:VMHYDRO_WATERCOURSE_DRAIN") %>%
  filter(DWITHIN(melbourne %>% sf::st_centroid(), distance = 10000, units = "meters")) %>%
  collect()

```

```{r paginate, eval = FALSE}
data <- vicmap_query(layer = "datavic:VMHYDRO_WATERCOURSE_DRAIN", wfs_version = "1.0.0") %>%
  collect()
```


```{r filter_spatial, eval = FALSE}
melbourne <- st_read(system.file("shapes/melbourne.geojson", package="VicmapR"))

vicmap_query(layer = "datavic:VMHYDRO_WATERCOURSE_DRAIN", CRS = 4326) %>%
#  filter(HIERARCHY == "L") %>%
  filter(BBOX(sf::st_bbox(melbourne))) %>%
  collect() %>%
  plot(max.plot = 2, lwd = 10, cex = 10)
```

```{r filter_spatial_2,eval=FALSE}
vicmap_query(layer = "datavic:VMHYDRO_WATERCOURSE_DRAIN", CRS = 4326, wfs_version = "1.0.0") %>%
  filter(VicmapR::DWITHIN(melbourne %>% sf::st_centroid(), distance = 1000, units = "meters")) %>%
  show_query()
  
  library(lineprof)

prof <- lineprof({vicmap_query(layer = "datavic:VMHYDRO_WATERCOURSE_DRAIN", CRS = 4326, wfs_version = "1.0.0") %>%
  filter(VicmapR::DWITHIN(melbourne %>% sf::st_centroid(), distance = 1000, units = "meters")) %>%
  collect()})

shine(prof)
#  filter(HIERARCHY == "L") %>%
vicmap_query(layer = "datavic:VMHYDRO_WATERCOURSE_DRAIN") %>%
  filter(INTERSECTS(melbourne %>% polygonFormat())) %>%
  show_query() %>% nchar()


# points <- list(melbz = c(144.9436,-37.81073))
# 
# closest_feature <- lapply(points, function(y) {
#   
# n_hits <- 0    
# i <- 0
# 
# while(n_hits < 1) { 
#   i <- i + 1
#   x <- 100*i
# data <- paste0("http://services.land.vic.gov.au/catalogue/publicproxy/guest/dv_geoserver/wfs?service=wfs&version=1.0.0&request=GetFeature&typeNames=datavic%3AVMHYDRO_WATERCOURSE_DRAIN&outputFormat=application%2Fjson&count=70000&srsName=EPSG%3A4326&CQL_FILTER=%28DWITHIN%28SHAPE%2C%20POINT%20%28",y[1],"%20", y[2],"%29%2C%20", x,"%2C%20meters%29%29") %>% sf::read_sf()
# n_hits <- nrow(data)
# 
# }
# point_poly <- sf::st_as_sf(x = data.frame(long = y[1], lat = y[2]),
#                            coords = c("long", "lat")) %>% sf::`st_crs<-`(4326)
# 
# data$distance_from_point <- sf::st_distance(point_poly, 
#                                             data, by_element = TRUE) 
# 
# data <- data %>%
#   dplyr::arrange(distance_from_point) %>%
#   dplyr::slice(1)
# 
# # nearest_points <- sf::st_nearest_points(point_poly, 
# #                                         data) %>% sf::st_cast("MULTIPOINT") %>%
# #   as.data.frame()
# 
# return(data)
# })

```


```{r loop_dwithin, eval=FALSE}
# remotes::install_github("justincally/VicmapR", ref = "3-vicmap-promise")
# library(sf)
# library(dplyr)
# library(VicmapR)

melbourne <- st_read(system.file("shapes/melbourne.geojson", package="VicmapR"))

some_point <- melbourne %>% sf::st_centroid()

# How many hits 
vicmap_query(layer = "datavic:VMHYDRO_WATERCOURSE_DRAIN", CRS = 4326, wfs_version = "2.0.0") %>%
  filter(VicmapR::DWITHIN(some_point, distance = 2000, units = "meters")) %>%
  feature_hits()

data <- vicmap_query(layer = "datavic:VMHYDRO_WATERCOURSE_DRAIN", CRS = 32755, wfs_version = "2.0.0") %>%
  filter(VicmapR::DWITHIN(some_point, distance = 1000, units = "meters")) %>%
  collect()
```


```{r query_select, eval = FALSE}
small.df <- vicmap_query(layer = "datavic:VMHYDRO_WATERCOURSE_DRAIN") %>%
  head(50) %>%
  select(HIERARCHY, PFI) %>%
  collect()
```
